Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-12-06 rnni_convexity.tex
typo
paper structure revised

\begin{abstract}
In this paper we study the graph of ranked phylogenetic trees where the adjacency relation is given by a local rearrangement of the tree structure.
Our work is motivated by tree inference algorithms, such as maximum likelihood and Markov Chain Monte Carlo methods, where the geometry of the search space plays a central role for efficiency and practicality of the optimisation.
We hence focus on understanding the geometry of the space (graph) of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
For example, we find exact estimates of the radius and diameter of the space.
Since the $\rnni$ graph is a generalisation of the classical nearest neighbour interchange ($\nni$) graph to ranked phylogenetic trees, we compare geometric and algorithmic properties of the two graphs.
Surprisingly, we discover that both geometric and algorithmic properties of $\rnni$ and $\nni$ are quite different.
For example, we establish convexity of certain natural subspaces in $\rnni$, which is not convex is $\nni$.
Furthermore, our results suggest that the complexity of computing distances in the two graphs is different.
\end{abstract}

@@

The nearest neighbour interchange ($\nni$) graph, defined on the set of phylogenetic trees with adjacency relation given by the interchange operation of two sister clades, has been known in mathematical biology literature for nearly 50 years \autocite{Robinson1971-ql,Moore1973-kk}.
Considered with the metric given by the length of a shortest path (graph-distance), this graph becomes a metric space.
Its geometry has been extensively studied \autocite{Dasgupta2000-xa, Li1996-zw, Gordon2013-fw, De_Jong2016-al}.
An important property of the $\nni$ graph is that computing distance is NP-hard \autocite{Dasgupta2000-xa}.
As a result no algorithm exists to compute the distance in practical time.
A consequence is that tree search and sampling algorithms pose a significant challenge even for moderately sized trees
\autocite{Whidden2016-kl}.

@@

The rest of this paper is organised as follows.
In Section~\ref{section:algorithms} we introduce three algorithms for exploring the $\rnni$ graph.
We use these algorithms in Section~\ref{section:geometry} to establish several geometrical properties of this graph, such as its diameter and radius.
Furthermore, we will prove in Section~\ref{section:caterpillar_convex} that distances can be computed in polynomial time in a subspace of the $\rnni$ space.
This result suggests that the complexity of computing distances in $\rnni$ is different from that in the classical $\nni$ space.
We will discuss this further in Section~\ref{section:cluster_theorem}, where we disprove a conjecture of \textcite{Gavryushkin2018-ol} and suggest its weaker version a plausible alternative.

>>>>

\begin{abstract}
In this paper we study the graph of ranked phylogenetic trees where the adjacency relation is given by a local rearrangement of the tree structure.
Our work is motivated by tree inference algorithms, such as maximum likelihood and Markov Chain Monte Carlo methods, where the geometry of the search space plays a central role for efficiency and practicality of the optimisation.
We hence focus on understanding the geometry of the space (graph) of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
For example, we find exact estimates of the radius and diameter of the space.
Since the $\rnni$ graph is a generalisation of the classical nearest neighbour interchange ($\nni$) graph to ranked phylogenetic trees, we compare geometric and algorithmic properties of the two graphs.
Surprisingly, we discover that both geometric and algorithmic properties of $\rnni$ and $\nni$ are quite different.
For example, we establish convexity of certain natural subspaces in $\rnni$, which are not convex is $\nni$.
Furthermore, our results suggest that the complexity of computing distances in the two graphs is different.
\end{abstract}

@@

The nearest neighbour interchange ($\nni$) graph, defined on the set of phylogenetic trees with adjacency relation given by the interchange operation of two sister clades, has been known in mathematical biology literature for nearly 50 years \autocite{Robinson1971-ql,Moore1973-kk}.
Considered with the metric given by the length of a shortest path (graph-distance), this graph becomes a metric space.
Its geometry has been extensively studied \autocite{Dasgupta2000-xa, Li1996-zw, Gordon2013-fw, De_Jong2016-al}.
An important property of the $\nni$ graph is that computing distances is NP-hard \autocite{Dasgupta2000-xa}.
As a result no algorithm exists to compute the distance in practical time.
A consequence is that tree search and sampling algorithms pose a significant challenge even for moderately sized trees
\autocite{Whidden2016-kl}.

@@

The rest of this paper is organised as follows.
In Section~\ref{section:algorithms} we introduce three algorithms for exploring the $\rnni$ graph.
We use these algorithms in Section~\ref{section:geometry} to establish several geometrical properties of this graph, such as its diameter and radius.
Furthermore, we will establish in Section~\ref{section:caterpillar_convex} that a certain subspace of $\rnni$ is convex and design a polynomial-time algorithm for computing $\rnni$-distances in that subspace.
This result suggests that the complexity of computing distances in $\rnni$ is different from that in the classical $\nni$ space.
We will discuss this further in Section~\ref{section:cluster_theorem}, where we disprove a conjecture of \textcite{Gavryushkin2018-ol} and suggest its weaker version a plausible alternative.
We will finish the paper with a discussion and open problems.
